---
title: "Collegiate Sports Budget Analysis"
subtitle: "STA/ISS 313 - Project 1"
author: "ggplot <3"
format: html
editor: visual
execute: 
  echo: false
  warning: false
---

```{r}
#| label: load data
sports <- read.csv("data/data.csv")
```

## Abstract

(1 paragraph): Project abstract.

------------------------------------------------------------------------

## Introduction

This data on collegiate sports budgets comes from Equity in Athletics Data Analysis. It has `r nrow(sports)` rows and `r ncol(sports)` columns, including numeric and categorical variables. The dataset has information on most colleges in the US, regardless of sports division, and consists of metrics such as revenue and expenditure for each sport at the school from 2015-2019. We chose the dataset because it has a large number of variables that will be interesting to analyze, and because, as college sports fanatics, we think the topic of collegiate sports budgets is interesting, relevant to us, and worthy of exploration. The variables we are primarily concerned with are `exp_men`, the school expenditure on men's sports, `rev_men`, revenue generated from men's sports for the school, `exp_women`, the school expenditure on women's sports, `rev_woman,` revenue generated from women's sports for the school, `total_exp_menwomen`, the total expenditure on men and women's teams for a certain sport, `year`, year of athletics, and `sector_name`, school type like "Private 4 year" for example.

```{r}
#| label: load packages
#| message: false

library(tidyverse)
library(patchwork)
library(janitor)
library(scales)
library(knitr)
```

## Question 1: Title that relates to the question you're answering

### Introduction

### Approach

After investigating the differences in revenue and expenditure for men and women in our first plot, we are also interested in investigating and analyzing how the relationship between profit and expenditure differ for men and women and whether the sport is basketball or other. To address this difference in relationship, we plan to create a separate scatter plot for men and women, colored by sports. We decided to use a scatter plot because we want to visualize and analyze the relationship between two quantitative variables. We also plan to add a best trend line through the data points to get a better visualization of the strength of relationship between profit and expenditure. Additionally, to further analyze how the relationship between profit and expenditure differ between men and women, we plan to calculate the correlation coefficient for each gender and compare them to see which correlation coefficient is higher.

### Analysis

```{r}
#| label: Q1-plot1-data-wrangling
#| message: false

mensrev <- sports %>%
  drop_na(rev_men) %>%
  group_by(year) %>%
  summarise(mean_rev_men = mean(rev_men))

womensrev <- sports %>%
  drop_na(rev_women) %>%
  group_by(year) %>%
  summarise(mean_rev_women = mean(rev_women))

mensexp <- sports %>%
  drop_na(exp_men) %>%
  group_by(year) %>%
  summarise(mean_exp_men = mean(exp_men))

womensexp <- sports %>%
  drop_na(exp_women) %>%
  group_by(year) %>%
  summarise(mean_exp_women = mean(exp_women))

plotdf <- mensrev %>%
  left_join(mensexp) %>%
  left_join(womensrev) %>%
  left_join(womensexp)

plotdf_pivoted <- plotdf %>% 
  pivot_longer(
    cols = c(mean_rev_men, mean_exp_men, mean_exp_women, mean_rev_women), 
    names_to = "category", 
    values_to = "amount"
  ) %>%
  mutate(gender = case_when(category == "mean_exp_men" | category == "mean_rev_men" ~ "Mens",
                            category == "mean_exp_women" | category == "mean_rev_women" ~ "Womens"),
         type = case_when(category == "mean_exp_men" | category == "mean_exp_women" ~ "Expenditure",
                            category == "mean_rev_men" | category == "mean_rev_women" ~ "Revenue"))

mensdata <- plotdf_pivoted %>%
  filter(gender == "Mens")
womensdata <- plotdf_pivoted %>%
  filter(gender == "Womens")
```

```{r}
#| label: Q1-plot1-graph
ggplot() +
  geom_point(data = plotdf_pivoted, aes(x = year, y = amount, shape = type, color = gender), size = 3) +
  geom_line(data = mensdata, aes(x = year, y = amount, group = year), linetype = "dashed", color = "#F8766D") +
  geom_line(data = womensdata, aes(x = year, y = amount, group = year), linetype = "dashed", 
            color = "#619CFF") +
  scale_y_continuous(labels = label_dollar(suffix = "K", scale = 1e-3)) +
  labs(title = "Average college sports revenue and expenditure",
       subtitle = "By gender",
       y = "",
       x = "Year",
       color = "Gender",
       shape = "Type") +
  theme_minimal()  +
  theme(legend.position = "top")
```

```{r}
#| label: Q1-plot2-data-wrangling
#| message: false

# filter out NA values
sports_mini = sports |>
  filter(!is.na(exp_men), !is.na(exp_women), !is.na(rev_men), !is.na(rev_women)) 

# pivot to create a gender column
sports_pivot <- sports_mini |>
  pivot_longer(
    cols = c("exp_men", "exp_women", "rev_men", "rev_women"),
    names_to = c("type_money", "gender"),
    values_to = c("amount"),
    names_sep = "_"
  )

# filter for only observations with year as 2015
sports_pivot <- sports_pivot |>
  filter(year == 2015) 

# pivot wider to create a separate column for revenue and expenditure
sports_pivot <- sports_pivot |>
  pivot_wider(
    names_from = type_money,
    values_from = amount
  )

# create a profit variable
sports_pivot <- sports_pivot |>
  mutate(
    profit = rev - exp
  )

# create separate data frame for only women
sports_pivot_women <- sports_pivot |>
  filter(gender == "women") 

# spotlight extreme outlier in women's data frame so we can remove it
sports_pivot_women |>
  filter(profit > 15000000) %>%
  select(year, institution_name, sports, profit) %>%
  kable(col.names = c("Year", "Institution", "Sport", "Profit"))

# mutate sports with values Basketball or Other
sports_pivot_women <- sports_pivot_women |>
  filter(profit < 15000000) |>
  mutate(
    sports =
      case_when(
        sports == "Basketball" ~ "Basketball",
        TRUE ~ "Other"
      )
  ) 

# create separate data frame for only men
sports_pivot_men <- sports_pivot |>
  filter(gender == "men") 

# mutate sports with values Basketball or Other
sports_pivot_men <- sports_pivot_men |>
  filter(profit < 15000000) |>
  mutate(
    sports =
      case_when(
        sports == "Basketball" ~ "Basketball",
        TRUE ~ "Other"
      )
  ) 

```

```{r}
#| label: Q1-plot2-correlation-coeff
# calculate the correlation coefficient for each gender and create a data frame
# to store the values
cors <- plyr::ddply(sports_pivot, c("gender"), summarise, cor = round(cor(exp, profit), 2))
```

```{r}
#| label: Q1-plot2-scatter-plots
#| message: false

# scatter plot for women
plot1 <- ggplot(sports_pivot_women, aes(x = exp, y = profit, color = sports)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "black") +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(
    title = "Relationship Between Profit and Expenditure for Basketball and Other Sports\nfor Women",
    subtitle = "in 2015",
    x = "Expenditure ($)",
    y = "Profit ($)",
    color = "Sport Type"
  ) +
  annotate(geom = "label", x = 0.5e+06, y = 3000000, label = "r = -0.57", size = 5,
           color = "black", fontface = 2) +
  annotate(geom = "text", x = 8e+06, y = 500000, label = "Profit = 0") +
  coord_cartesian(clip = "off", ylim = c(-5100000, 4000000)) +
  scale_y_continuous(breaks = c(-5000000, -2500000, 0, 2500000),
                     labels = c("$-5M", "$-2.5M", "$0M", "$2.5M")) +
  scale_x_continuous(breaks = c(0e+00, 2e+06, 4e+06, 6e+06),
                     labels = c("$0M", "$2M", "$4M", "$6M")) +
  scale_color_manual(values = c("#8E7DBE", "#D88C9A")) +
  theme_minimal() +
  theme(
    legend.position = c(0.9, 0.9),
    legend.box.background = element_rect(fill = "white", color = "black")
  )

# scatter plot for men
plot2 <- ggplot(sports_pivot_men, aes(x = exp, y = profit, color = sports)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "black") +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(
    title = "Relationship Between Profit and Expenditure for Basketball and Other Sports\nfor Men",
    subtitle = "in 2015",
    x = "Expenditure ($)",
    y = "Profit ($)",
    color = "Sport Type"
  ) +
  annotate(geom = "label", x = 1.2e+06, y = 2.5e+07, label = "r = 0.54", size = 5,
           color = "black", fontface = 2) +
  annotate(geom = "text", x = 2e+07, y = 1.5e+06, label = "Profit = 0") +
  coord_cartesian(clip = "off", ylim = c(-5000000, 30000000)) +
  scale_x_continuous(breaks = c(0.0+00, 5.0e+06, 1.0e+07, 1.5e+07), 
                     labels = c("$0M", "$5M", "$10M", "$15M")) +
  scale_y_continuous(breaks = c(0.0+00, 1e+07, 2e+07, 3e+07, 4e+07), 
                     labels = c("$0M", "$10M", "$20M", "$30M", "$40M")) +
  scale_color_manual(values = c("#8E7DBE", "#D88C9A")) +
  theme_minimal() +
  theme(
    legend.position = c(0.9, 0.9),
    legend.box.background = element_rect(fill = "white", color = "black")
  )

plot1
plot2
```

### Discussion

This first chart reveals several key finding about how spending and revenue differ for men's and women's college sports. First, men's sports are, on average both spending and earning far more than women's sports. Men's sports earn on average about x more than women's sports, and spend about y more. This gap was persistent from 2015 to 2019, in fact even slightly growing across this time period. We were surprised to see that the gap was this dramatic, especially given Title IX spending rules; we were also surprised that the gap grew during a time period where so many people have been focused on equity in sports. A second finding evident in this chart is that men's sports, on average, earn more than they spend, while the opposite is true for women's sports. This positive gap in earning is also greater in men's sports than the respective negative gap in earning for women's, a finding which is also consistent across years.

From the two graphs, we can see that the relationship between profit and expenditure differs greatly for men and women. For men, there appears to be a moderate positive relationship between profit and expenditure while for women there appears to be a moderate negative relationship between profit and expenditure. This trend is probably because although colleges spend millions of dollars in funding for women collegiate sport programs, these women collegiate sport programs do not generate enough revenue for positive profit. This is most likely due to women collegiate programs obtaining less publicity in corporate sponsorship and media viewership, according to the NPR article on unequal profits. It is reasonably expected that the relationship between profit and expenditure is positive as shown in men's collegiate sports programs; however, this is not the case for women collegiate sports programs since external uncontrollable factors such as media viewership and corporate sponsorship influence how much revenue and thus profit is generated from advertisement, ticket sales, and sponsorships.

The magnitude of the r-squared value 0.57 for women is slightly greater than 0.54 for men. This means that the relationship for profit and expenditure for women follow more of a linear relationship than for men. This can be due to women's collegiate sports often times displaying a consistent trend in generating low profits for an increase in expenditure because of the collegiate sports' industry of underlying gender bias and discrimination. This relationship for men is less linearly correlated and shows more spread because different schools and divisions will have more promising teams than other schools and divisions which influences the profit generated from expenditures.

Interestingly, the relationship between profit and expenditure for basketball is positive and stronger than the relationship between profit and expenditure for other sports where the relationship is negative. This makes sense because men's collegiate basketball is one of the most popular collegiate sport that generates enormous amounts of publicity and profit. However, for women, no matter the type of sport, the relationship between profit and expenditure is negative and similar. This similarity in trend for all sports for women shows that there might be some underlying gender discrimination and under-representation in terms of profitable opportunities in collegiate sports for women.

## Question 2: Differences in sports expenditure between public and private universities

## Introduction

We are interested in seeing how the mean expenditure per sport of Division I-FBS schools from 2015-19 differs for public and private universities. The Division I-FBS is the top division in college sports that at the time of our data being collected, consisted of 126 universities, 18 private and 108 public. The parts of our data that are necessary to answer this question are the sport, whether or not a Division I-FBS school is public or private, and difference in the mean expenditure for public and private schools over the years for each respective sport.

We are interested in this question as in general, private schools have a larger endowment than public schools, meaning they have more money to spend on athletics. On the other hand, public schools are larger, have a larger fan base, consistently outperform private schools in football, and receive state subsidized funding allowing them to possibly spend more on sports. With both schools having a possible advantage in expenditure spending, we are interested to see which schools spend more money on their athletics programs.

### Approach

The first plot we are going to make to answer this question is a bidirectional bar chart plotting the difference in mean expenditure for public and private schools for for each respective sport. We plan to have sport on the y-axis and the difference in expenditure on the x-axis. A bidirectional bar chart is the best chart to show the difference in mean expenditure per sport as it will clearly show which sports private schools spend more money on and which sports public schools spend more money on. It will clearly show this relationship as the bars will point in opposite directions depending on whether private or public schools spend more. Furthermore, this chart will also by how much the mean expenditure is different between the private and public schools for each sport, giving us a good indication on whether or not there are simply minor or major spending differences. This chart will also allow us to see the general trend in spending differences and the range that spending differences gernally fall between.

Public universities tend to outperform private universities in football games. Nevertheless, based on the initial visualization, it was discovered that private universities allocate more funds for this sport than their public counterparts. Consequently, we conducted further analysis on the data distribution of the two major university sports, namely basketball and football. These sports were chosen due to their large number of data points. To display the distributions, we employed box plots that were facet-wrapped by sports and color-coded according to the type of university, whether public or private.

### Analysis

```{r}
#| label: Q2-part-1-vis
#load data 
sports <- read.csv("data/data.csv")

#filter to get d1-FBS schools
class_sports<- sports |> 
  filter(classification_code== 1) 

#group based on sport and whtehr or not private or public and calculate the mean expenditure over the 5 years for each sport 
d1_priv_pub <- class_sports|> 
  group_by(sports, sector_name) |> 
  mutate(mean_exp = mean(total_exp_menwomen, na.rm = TRUE)) |> 
  ungroup()

#get distinct columns for each sport and clean data set setting the mean expenditure to negative for the direciton of the bidirectional bar chart
priv_pub <- d1_priv_pub |>  
  filter(year == 2015) |>  
  select(sports, mean_exp, sector_name) |> 
  distinct() |> 
  mutate(mean_exp = ifelse(sector_name == "Public, 4-year or above", -mean_exp,
                           mean_exp)) 
#create a vector of the sports I want to remove from data 
rem <- c("Track and Field, Indoor", "Squash", "Track and Field, X-Country",
         "Track and Field, Outdoor", "Other Sports",
         "Synchronized Swimming", "Diving",
         "Bowling", "Sailing", "Skiing", "Rifle", "Equestrian")

#clean and pivot wider code to get single column for sports with columns for the mean expenditre and difference for public and private school mean expenditure
pivot_priv <- priv_pub |> 
  select(mean_exp, sector_name, sports) |>  
  pivot_wider(names_from = sector_name,
              values_from = mean_exp) |> 
  clean_names() |>  
  mutate(difference = public_4_year_or_above + 
           private_nonprofit_4_year_or_above,
         difference_abs = abs(difference)) |> 
  mutate(out_spend = ifelse(difference >0, "Private", "Public")) %>% 
  filter(!(sports %in% rem))

#create visualizaton of bidirectionla bar chart
ggplot(pivot_priv, aes(x = difference, y = fct_reorder(sports, difference), 
                       fill = out_spend))+
  geom_bar(stat= "identity")+
    scale_x_continuous(labels = dollar_format(scale = 1e-6, prefix = "$", 
                                            suffix = "M"))+
  labs(x = "Difference in Mean Expenditure",
       y = "Sport",
       fill = "Bigger Spender?",
       title = "Bidirectional Bar Chart of Difference in Mean Expenditure 
       for Public and Private D1-FBS Schools")+
  scale_fill_viridis_d()+
  theme_minimal()
```

```{r}
#| label: Question-2-Part-2-data-wrangling
q2_2 <- sports |>
  filter(classification_code == 1 & (sports == "Basketball" | sports == "Football") ) |> 
  select(sports, sector_name, total_exp_menwomen) |> 
  #renaming sector name values, so thatt it will look nice on the graph
  mutate(sector_name = recode(sector_name, "Private nonprofit, 4-year or above" = "Private", "Public, 4-year or above" = "Public")) 

```

```{r}
#| label: Question-2-Part-2-data-visualization

ggplot(
  q2_2, 
  aes(y = total_exp_menwomen, color = sector_name)
) +
  #creating a boxplot
  geom_boxplot() + 
  #fixing the y axis labels
  scale_y_continuous(labels = label_number(prefix = "$", suffix = "M", scale = 1e-6)) + 
  #facet wrapping by sports
  facet_wrap(~ sports, strip.position = "bottom") + 
  theme_minimal() +
  labs( 
    y = "Total expenditure",
    title = "Total Expenditures of Private and Public Universities", 
    subtitle = "On Basketball and Football in 2015-2019"
  ) +
  theme(
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.2, 0.8),
    legend.box.background = element_rect(fill = "white",
                                         color = "white"),
    legend.title = element_blank()
  ) 
```

### Discussion

Private schools outspent public schools on every sport except for swimming, fencing, and wrestling, only by around \$500,000 or less. The biggest surprise to us was that the private schools spend around \$4.75 million more than public universities on their football programs. A private school has not won the college football national championship in 20 years, so it is very surprising that over the years from 2015-19 the mean football expenditure across private schools was \$4.75 million more than that of public schools. Other than Football and Basketball, the general trend in the data is that the mean expenditure of private and public universities is within \$1 million of each other. The data probably looks the way it does as since private schools are generally not as good as public schools at football, they spend more money on their programs for things like recruiting and facilities in order to entice recruits to come without the storied success schools like Alabama have. Private schools may spend more money on basketball simply due to the fact that public schools are more concerned with their football programs and do not care to spend as much money on basketball. When dealing with budgets that are in the multi million dollar range, a couple hundred thousand dollars is not that large of a number in the bigger picture, that is probably why most of the sports difference in mean expenditure falls between \$1 million.

The second visualization indicates that private universities spend more on football than public universities, with a noticeable difference in median expenditure. This implies that half of all public universities allocate less funding to football compared to half of all private universities. However, there are multiple exceptions within the public universities, as some spend more than private universities, which may be attributed to their success in football. Despite the significant difference in the first plot, we found out from this visualization that there are some public universities that spend more on football than private ones. On the other hand, for basketball, there is no significant difference in median expenditure between public and private universities. Nonetheless, public universities have a bit wider range of expenditure on basketball than private universities. Overall, universities tend to allocate more funding to football than basketball.
