---
title: "Collegiate Sports Budget Analysis"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

```{r load-data}
sports <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-29/sports.csv')
```

This data on collegiate sports budgets comes from Equity in Athletics Data Analysis. It has `r nrow(sports)` rows and `r ncol(sports)` columns, including numeric and categorical variables. We chose the dataset because it has a large number of variables that will be interesting to analyze, and because we think the topic of collegiate sports budgets is interesting and worthy of exploration.

# Questions

The two questions you want to answer.

1.  How did the relationship between expenditure and participation of collegiate sports change over time for different genders?

2.  Which major sports of different type of colleges have equal revenue for men and women?

    Analysis plan

A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
