---
title: "Collegiate Sports Budget Analysis"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| warning: false
#| message: false

library(tidyverse)
```

# Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

```{r load-data}
#| message: false
sports <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-29/sports.csv')
```

This data on collegiate sports budgets comes from Equity in Athletics Data Analysis. It has `r nrow(sports)` rows and `r ncol(sports)` columns, including numeric and categorical variables. We chose the dataset because it has a large number of variables that will be interesting to analyze, and because we are college sports fanatics, we think the topic of collegiate sports budgets is interesting, relevant to us, and worthy of exploration.

# Questions

The two questions you want to answer.

1.  How did the relationship between expenditure and revenue of collegiate sports change over time for different genders?

2.  How does the total expenditure for both men and women differ by each year and how does this relationship differ by public or private sectors for 4 year or above colleges?

# Analysis

#### Question 1

We want to investigate how the relationship between college expenditure on collegiate sports and revenue generated from collegiate sports changed over the years for men and women. We plan to implement two time series plots to investigate this relationship. For our first plot, we plan to plot the variable exp_men on the x-axis and the variable rev_men on the y-axis to create a scatter plot. Then we plan to facet wrap by the variable year so we can observe how the relationship between college expenditure on collegiate sports and revenue generated from collegiate sports differs by year. Similarly, for our second scatter plot, we plan to plot the variable exp_women on the x-axis and the variable rev_women on the y-axis. We will also facet wrap by year and observe the differences in this relationship by year.

#### Question 2

For our second question, we plan to investigate how total expenditure for both men and women combined differs for each year from 2015 to 2019. We will use a box plot with the year variable plotted on the x-axis as a factor and the total_exp_menwomen variable plotted on the y-axis as a numeric. Because our data from TidyTuesday is too large with over 130,000 observations, we will subset our data by deciding which sports we want to specifically investigate how total expenditure for men and women differ by year. We are interested in investigating whether total expenditure by year differs also by the type of 4 year or above sectors: private and public. So, we will filter for only sectors that are "Public, 4-year or above" and "Private nonprofit, 4-year or above". Then, we plan to facet wrap by sector as an additional layer to our boxplot so we can see the differences in the distribution of total expenditure over the years by the two different sectors.

```{r}
#| label: Question-2-filtering-data
sports = sports |>
  filter(sector_name == "Private nonprofit, 4-year or above" | 
           sector_name == "Private nonprofit, 4-year or above")
```

The code above is an example of how we plan to filter our data for only those observations with sector name as "Private nonprofit, 4-year or above" or "Private nonprofit, 4-year or above".
